Demo: PhoneLets -- Offloading the Phone off your Phone for Energy, Cost and Network Load Optimization

### 2014-07-15 paper.tex
hopefully camera-ready

\usepackage{enumitem}
% \setlist{nolistsep}

@@

\clubpenalty=10000 
\widowpenalty = 10000 
\newfont{\mycrnotice}{ptmr8t at 7pt}
\newfont{\myconfname}{ptmri8t at 7pt}
\let\crnotice\mycrnotice%
\let\confname\myconfname%
\permission{Permission to make digital or hard copies of part or all of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage, and that copies bear this notice and the full citation on the first page. Copyrights for third-party components of this work must be honored. For all other uses, contact the owner/author(s). Copyright is held by the author/owner(s).}
\conferenceinfo{MobiCom'14,}{September 7-11, 2014, Maui, Hawaii, USA.} 
\copyrightetc{ACM \the\acmcopyr}
\crdata{ACM 978-1-4503-2783-1/14/09. \\
http://dx.doi.org/10.1145/2639108.2641744}

@@

% end author
\maketitle
% Use the following at camera-ready time to suppress page numbers.
% Comment it out when you first submit the paper for review.
%\thispagestyle{empty}
\begin{abstract}
This demo presents how phone functionality can be offloaded from a smartphone over wireless link to a \emph{PhoneLet} by sharing one SIM card across multiple devices. This can lead to significant cost and network load reductions by reducing the number of simultaneously connected mobile clients. Furthermore, it can bring significant energy reductions for the mobile user when connected to a powered \emph{PhoneLet}. It absorbs the energy cost of inefficient mobile applications' communication patterns, instead providing connectivity for the user over a WiFi link.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=0.9\columnwidth]{figs/arch}
\caption{Architecture of PhoneLets and interaction with different types of servers}
\label{fig:arch}
\end{figure}

The overall concept and system architecture is shown in Fig.~\ref{fig:arch}. There are three types of entities: 1) mobile server, a mobile device which can share its SIM with other clients, 2) SIM server which may host and provide more sets of credentials over local or wide-area networks and 3) a mobile client which can use either of the methods.

The SIM card is a special I/O device in a mobile platform: due to its security requirements it is not exposed to the platform's operating system. Instead, it is managed by the modem's firmware which routes all communication between the network and the subscriber module. Standard remote IO approaches are similar in their nature (\eg RIO~\cite{Sani:2014hs} enables transparently accessing I/O devices of a remote mobile system) but are not applicable in this setting.

Interaction with the subscriber module requires lower level protocols understood by the modem firmware. Our design extends the standard RSAP (Remote SIM Access Protocol)~\cite{Anonymous:2008vz} by using a SIM server which is not capable of mobile connectivity itself and by using arbitrary communication links to access a remote SIM card (\emph{no SIM} in the mobile client). Our work for this demo focused primarily on the client and SIM server design, and we are using off-the-shelf hardware as the mobile server to demonstrate backwards-compatibility.

\begin{figure}[t!]
\centering
\includegraphics[width=0.95\columnwidth]{figs/sequence}
\caption{Communication sequence between client, server and remote SIM card}
\label{fig:sequence}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.95\columnwidth]{figs/client}
\caption{Client prototype}
\label{fig:client}
\end{figure}

@@

Energy consumption of a Samsung Galaxy S4 was measured\footnote{Using Monsoon Power Monitor \url{http://www.msoon.com/LabEquipment/PowerMonitor/}} in a number of scenarios. The different values are presented in Table.~\ref{tab:energy}\footnote{Current drain (mA) is chosen over power (mW) as power depends on voltage that decays with the battery discharge}.

@@

The default BT implementation of RSAP is not efficient enough - maintaining access alone consumes almost three times the amount of energy of an idle device connected to a 3G network. Instead, we have implemented SIM sharing over WiFi using our custom client and verified that the BT version is unnecessarily wasteful - once the PhoneLet has registered to the network, there is no communication with the SIM server for as long as the device is idle. For example, a phonecall is successfully received after 30 minutes of no activity. The current Android platform does not allow RSAP to run over WiFi, but based on our measurements we expect energy consumption of an optimized implementation to be very close to that of an idle WiFi channel - 3.5x less than BT.

Finally, we measured energy consumption of WiFi tethering. With just one idle client it is almost 3x higher than the BT case. RSAP is effectively an alternative to WiFi tethering when the mobile client provides connectivity to other devices. So similarly to above, we expect energy consumption of an optimized WiFi mobile client to be close to 10x less than WiFi tethering.

These are early results, but they point to a potential use-case where a powered \emph{PhoneLet} is registered to the network on behalf of a WiFi-connected client at a considerable energy benefit. Outside the scope of this demo, the benefit would be further improved when network access is shared across multiple such devices and SIM access load is balanced across them as the savings for each individual device would add up.

@@

One cellular network in UK has offered a service to make phonecalls from tablets and laptops over WiFi to its users using a special app\footnote{O2 TU Go \url{http://www.o2.co.uk/apps/tu-go}} further blurring the line between mobile and the Internet. If WiFi Internet connectivity is available, the approach could help reduce energy consumption by disabling cellular radio.

@@

Reprogrammable SIM cards~\cite{OFcom:2012tx} such as embedded SIM~\cite{Association:2013ub} have been proposed for provisioning in M2M communication. They maintain a physical secure element to store credentials, but provide mechanisms to manage them over the air. The architecture, allows operators to retain full control over the credentials, including when new ones can be installed or old ones removed, with necessarily centralized management and reduced flexibility on the client side.

>>>>

\usepackage{enumitem}
\setlist{nolistsep}

@@

\newfont{\mycrnotice}{ptmr8t at 7pt}
\newfont{\myconfname}{ptmri8t at 7pt}
\let\crnotice\mycrnotice%
\let\confname\myconfname%
\permission{Permission to make digital or hard copies of part or all of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage, and that copies bear this notice and the full citation on the first page. Copyrights for third-party components of this work must be honored. For all other uses, contact the owner/author(s). Copyright is held by the author/owner(s).}
\conferenceinfo{MobiCom'14,}{September 7-11, 2014, Maui, Hawaii, USA.} 
\copyrightetc{ACM \the\acmcopyr}
\crdata{978-1-4503-2783-1/14/09. \\
http://dx.doi.org/10.1145/2639108.2641744}
\clubpenalty=10000 
\widowpenalty = 10000

@@

% end author
\maketitle
% Use the following at camera-ready time to suppress page numbers.
% Comment it out when you first submit the paper for review.
%\thispagestyle{empty}
\begin{abstract}
This demo presents how phone functionality can be offloaded from a smartphone over wireless link to a \emph{PhoneLet} by sharing one SIM card across multiple devices. This can lead to significant cost and network load reductions by decreasing the number of simultaneously connected mobile clients. Furthermore, it can save energy for the mobile user when connected to a powered \emph{PhoneLet} by offloading phone functionality. It absorbs the energy cost of online presence and inefficient mobile applications' communication patterns, instead providing connectivity for the user over a WiFi link.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figs/arch}
\caption{Architecture of PhoneLets and interaction with different types of servers}
\label{fig:arch}
\end{figure}

Figure~\ref{fig:arch} shows the overall concept and system architecture. There are three types of entities: 1) mobile server, a mobile device which can share its SIM with other clients, 2) SIM server which may host and provide more sets of credentials over local or wide-area networks and 3) a mobile client which can use either of the methods.

The SIM card is a special I/O device in a mobile platform: due to its security requirements it is not exposed to the platform's operating system. Instead, it is managed by the modem's firmware which routes all communication between the network and the subscriber module. Standard remote IO approaches are similar in their nature (\eg RIO~\cite{Sani:2014hs} enables transparently accessing I/O devices of a remote mobile system) but are not applicable in this setting as they require OS support.

Interaction with the subscriber module requires lower level protocols understood by the modem firmware. Our design extends the standard RSAP (Remote SIM Access Protocol)~\cite{Anonymous:2008vz} by using a SIM server which is not capable of mobile connectivity itself and by using arbitrary communication links to access a remote SIM card (\emph{no SIM} in the mobile client). Our work for this demo focused primarily on the client and SIM server design. We use off-the-shelf hardware as the mobile server for backwards-compatibility.

\begin{figure}[t!]
\centering
\includegraphics[width=\columnwidth]{figs/sequence}
\caption{Communication sequence between client, server and remote SIM card}
\label{fig:sequence}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=\columnwidth]{figs/client}
\caption{Client prototype}
\label{fig:client}
\end{figure}

@@

We measured energy consumption of a Samsung Galaxy S4\footnote{Using Monsoon Power Monitor \url{http://www.msoon.com/LabEquipment/PowerMonitor/}} in a number of scenarios. Table.~\ref{tab:energy}\footnote{Current drain (mA) is chosen over power (mW) as power depends on voltage that decays with the battery discharge} presents energy consumption of the different device radios, depending on their use, ranging from an idle device with only a particular radio enabled, to different methods of sharing connectivity and simple application usage patterns.

@@

The default BT implementation of RSAP is not efficient enough - maintaining access alone consumes almost three times the amount of energy of an idle device connected to a 3G network. Instead, we have implemented SIM sharing over WiFi using our custom client and verified that the BT version is unnecessarily wasteful - once the PhoneLet has registered to the network, there is no communication with the SIM server for as long as the device is idle. For example, a phone call is successfully received after 30 minutes of no activity. The current Android platform does not allow RSAP to run over WiFi, but based on our measurements we expect energy consumption of an optimized implementation to be close to that of an idle WiFi channel - 3.5x less than BT.

Finally, we measured energy consumption of WiFi tethering. With just one idle client it is almost 3x higher than the BT case. RSAP is effectively an alternative to WiFi tethering when the mobile client provides connectivity to other devices. Therefore we expect energy consumption of an optimized WiFi mobile client to be close to 10x less than WiFi tethering.

These are early results, but they point to a potential use-case where a powered \emph{PhoneLet} is registered to the network on behalf of a WiFi-connected client at a considerable energy benefit. The benefit would be further improved when network access is shared across multiple such devices and SIM access load is balanced across them as the savings for each individual device would add up. This use-case is outside the scope of the current demo.

@@

One cellular network in UK has offered a service to make phonecalls from tablets and laptops over WiFi to its users using a special application\footnote{O2 TU Go \url{http://www.o2.co.uk/apps/tu-go}}, blurring the line between mobile and the Internet. If WiFi Internet connectivity is available, the approach could help reduce energy consumption by disabling cellular radio.

@@

Reprogrammable SIM cards~\cite{OFcom:2012tx} such as embedded SIM~\cite{Association:2013ub} have been proposed for provisioning in M2M communication. They maintain a physical secure element to store credentials, but provide mechanisms to manage them over the air. The architecture allows operators to retain full control over the credentials, including when new ones can be installed or old ones removed through centralized management and subsequently reduced flexibility on the client side.
